```         
```

---
title: "My project"
format: html
editor: visual
---

## Project description

The data covers the status of water, sanitation, and hygiene services in 39 health care facilities in four districts of the Upper East Region in Ghana.\
My goal is to analyse this to assess the availability, functionality, and quality of water, sanitation, and hygiene (WASH) services across the 39 health care facilities, identify facility-level and district-level gaps, and develop data-driven recommendations for improving WASH service delivery and compliance with national and WHO standards.

## Import

```{r}
library(tidyverse)
library(here)

```

```{r}
raw_data <- read_csv(here::here("data/raw/hcfs_assessment.csv"), locale = locale(encoding = "ISO-8859-1"))

```

```{r}
processed_data <- raw_data |> 
  rename(
  `toilet_light_vent` = `Sufficient lighting and ventilation in toilets`,
    `cleaning_recs` = `No sufficient lighting and ventilation in toilets`, 'hcf_name' = 'Health Care Facility name', 'hcf_loc' = 'Health Care Facility location', 'hcf_level' = 'Level of health care facility', 'hcf_ownwership' = 'Facility ownership', 'hcf_ownership_det' = 'Facility ownership_Other', 'refugee_centre' = 'Refugee centre near facility', 'doctors_total' = 'Total number of doctors', 'doctors_female' = 'Number of female doctors', 'doctors_male' = 'Number of male doctors', 'nurses_total' = 'Total number of nurses', 'nurses_female' = 'Number of female nurse', 'nurses_male' = 'Number of male nurses', 'cleaners_total' = 'Total number of cleaners', 'cleaners_female' = 'Number of female cleaners', 'cleaners_male' = 'Number of male cleaners', 'staff_other' = 'Number of other staff', 'staff_other_female' = 'Number of female other staff', 'staff_other_male' = 'Number of male other staff', 'staff_comments' = 'Comments on number of staff', 'number_outpatients' = 'Number of outpatients', 'outpatient_unit' = 'OPD units', 'number_inpatients' = 'Number of inpatients', 'inpatient_units' = 'inpatients unit', 'patients_comments' = 'Number of patients comments', 'services' = 'Services offered', 'services_other' = 'Other services', 'water_supply_main' = 'Main water supply', 'water_supply_main_loc' = 'Main water supply location', 'water_supply_main_avl' = 'Water availability from main supply at time of survey', 'water_supply_comments' = 'Additional comments on water supply', 'toilet_type_patients' = 'Common types of toilets for patients', 'toilet_type_other' = 'Other toilet types', 'toilet_type_comments' = 'Comments on toilet', 'toilet_usable' = 'At least one toilet usable at survey', 'hyg_station_funct' = 'Functioning hygiene stations at least 5 m from toilet', 'toilet_sep_staff' = 'Separate toilets for staff', 'toilet_AYFS' = 'Toilet in the AYFS', 'toilet_sep_male_female')
  
  
```
